Quantity discrimination, decision-making, and the role of early-life conditions in a lizard

### 2025-09-29 ms.qmd
updates

Here, we investigated quantity discrimination in the common garden skink (_Lampropholis guichenoti_) using a spontaneous choice test where lizards were presented with different numbers of food items. Our aim was to understand if this species could use the OFS to discriminate between small quantities of prey in a spontaneous choice test. In addition, we examined how prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature can affect quantity discrimination and decision-making in foraging contexts. To do so, we manipulated CORT concentration (CORT-treated or a sham control) in _L. guichenoti_ eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3ºC or Hot - 28 ± 3ºC) in a 2x2 factorial design. Then, we assessed numerical discrimination abilities on five tests that differed in the relative number of food items: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets. Previous studies showed that skink species can discriminate between different quantities of food using both the ANS and the OFS [@szabo_spontaneous_2021; @szabo2024spontaneous]. We predicted _L. guichenoti_ to be able to discriminate between different numbers of crickets even when controlling for other continuous variables like total length and area (i.e. by using the OFS). However, we predicted that numerical discrimination would be affected by early-life conditions. Specifically, we hypothesized that lizards incubated at higher temperatures would outperform those incubated at lower temperatures in the numerical discrimination task as previous studies have reported enhanced cognitive abilities ectotherms incubated under warmer conditions [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_quantity_2019]. Furthermore, we predicted that increased CORT levels would impair numerical discrimination [@szuran_water_1994; @zhu_prenatal_2004], with more pronounced effects at lower temperatures.

@@

*Egg collection and incubation* -- Between mid-October 2023 to the end of February 2024, we placed small boxes (12.5 L x 8.3 W x 5 H cm) containing vermiculite on one side of the communal enclosures (see above) to provide females with a place to lay the eggs. We checked for eggs in the boxes three days a week. Eggs were assigned a clutch number and an individual identity. Then, we treated eggs with CORT or a vehicle control (see _Manipulating Early Thermal and CORT Environments_ details below)and placed each egg in individual cups (80 mL) with moist vermiculite (12 g water to 4 g vermiculite) and covered the cups with plastic wrap to retain moisture. Each egg was incubated individually in incubators (LATWIT 2X5D-R1160) programmed to two different thermal regimes (see _Manipulating Early Thermal and CORT Environments_ details below) until hatching. Incubators were checked for the presence of hatchlings three times a week. Egg and hatchling mortality was not higher than 13% in any of the treatments (Leibold et al., In prep.).

>>>>

Here, we investigated quantity discrimination in the common garden skink (_Lampropholis guichenoti_) using a spontaneous choice test where lizards were presented with different numbers of food items. Our aim was to understand if this species could use the OFS to discriminate between small quantities of prey in a spontaneous choice test. In addition, we examined how prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature can affect quantity discrimination and decision-making in foraging contexts. To do so, we manipulated CORT concentration (CORT-treated or a sham control) in _L. guichenoti_ eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3ºC or Hot - 28 ± 3ºC) in a 2x2 factorial design. Then, we assessed numerical discrimination abilities on five tests that differed in the relative number of food items: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets. Previous studies showed that skink species can discriminate between different quantities of food using both the ANS and the OFS [@szabo_spontaneous_2021; @szabo2024spontaneous]. We predicted _L. guichenoti_ to be able to discriminate between different numbers of crickets even when controlling for other continuous variables like total length and area (i.e. by using the OFS). However, we predicted that numerical discrimination would be affected by early-life conditions. Specifically, we hypothesized that lizards incubated at higher temperatures would outperform those incubated at lower temperatures in the numerical discrimination task as previous studies have reported enhanced cognitive abilities in ectotherms incubated under warmer conditions [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_quantity_2019]. Furthermore, we predicted that increased CORT levels would impair numerical discrimination [@szuran_water_1994; @zhu_prenatal_2004], with more pronounced effects at lower temperatures.

@@

*Egg collection and incubation* -- Between mid-October 2023 to the end of February 2024, we placed small boxes (12.5 L x 8.3 W x 5 H cm) containing vermiculite on one side of the communal enclosures (see above) to provide females with a place to lay the eggs. We checked for eggs in the boxes three days a week. Eggs were assigned a clutch number and an individual identity. Then, we treated eggs with CORT or a vehicle control (see _Manipulating Early Thermal and CORT Environments_ details below) and placed each egg in individual cups (80 mL) with moist vermiculite (12 g water to 4 g vermiculite) and covered the cups with plastic wrap to retain moisture. Each egg was incubated individually in incubators (LATWIT 2X5D-R1160) programmed to two different thermal regimes (see _Manipulating Early Thermal and CORT Environments_ details below) until hatching. Incubators were checked for the presence of hatchlings three times a week. Egg and hatchling mortality was not higher than 13% in any of the treatments (Leibold et al., In prep.).
